Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-05 Dail_Mad_ext.tex
lunch

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in
abundance---hypotheses relating to the importance and existence of
phenomenon such as density-dependent population regulation, population
cycling, and spatial synchrony
\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999}.
In applied contexts, population models are used for
estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004}.
In order to address these questions, several complicating factors must
be confronted when fitting population models to data.
First, deterministic models of population
dynamics are virtually always inadequate due to process variation,
the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001}. %clark_bjornstad:2004}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.
%Failure to account for process
%variation and observation error can bias the estimators of abundance
%and related parameters.

State-space models are a widely used approach for
studying population dynamics while accounting for process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Conventional state-space
models are simply time-series models in which the true state of the
system (e.g. population size during each year) is observed
imperfectly. One reason for the widespread adoption of state-space
models in ecology is that failure to account for process variation and
observation error can bias estimators of abundance and population
growth parameters \citep{kery_etal:2009}. For instance, the strength of
density dependence will be over-estimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.
%Originally developed for xxx (cite Kalman filter papers),
%The ability to model both the ecological state process and
%the observation process has made them relevant in numerous ecological
%applications [need to clean this up]. State-space models have been
%defined differently by different authors, but we follow
%\citet{buckland_etal:2004} the definition : ``xxx'' .

One of the simplest state-space models can be described as follows.
% In their simplest form, state-space models include a Markovian model
%for the ecological process of interestTypically, the
%ecological
%process is described by a Markovian model that includes a
%deterministic description of population change as well as two random
%sources of variation for ``process variation''. The model for the
%Robservation process is typically phenomenological in the sense that
%observation error is not defined explicitly. As an example,
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model
(e.g. exponential, logistic, etc \dots) and stochastic model allowing
for random process variation, which we denote $\grave{o}$ . As an
example, under exponential population growth, the model could be
written as:
\begin{gather}
N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
X_t = N_t + \zeta_t
\end{gather}
where $r$ is the intrinsic rate of increase. Typically, both random
effects are assumed to follow normal distributions: $\grave{o}_t \sim
\mathrm{N}(0, \sigma)$ and $\zeta_t \sim \mathrm{N}(0, \tau)$
\citep{devalpine_hastings:2002,dennis_etal:2006}. Even though this is
the most widespread approach for modeling population dynamics using
time-series data, several problems are immediately evident. First, if
abundance is a non-negative integer, as it always is, there is clearly
a problem with this formulation because the normally distributed
random effects violate this constraint. To avoid this issue,
ecologists often replace $N_t$ with $\log(Y_t/A)$   where $A$ is the
area surveyed---i.e.,
abundance is replaced by the natural log of population density, which
is assumed to normally distributed. This is desirable because this
allows for the fitting of models assuming normally distributed
residuals, which are computationally much less intensive than most
alternatives. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with
some small, but positive number, although the effect of this is rarely
discussed.

The problems mentioned thus far are minor in comparison to the more
serious issue that the parameters of the model are often not
identifiable. Specifically, the Markovian nature of the model implies
the following likelihood
\[
LIKELIHOOD
\]
The first term is the probability density for the first observation in
the time series, $Y_t$, yet it is intuitively not possible to estimate the
parameter(s) of this distribution using only a single
observation. This problem has resulted in several papers with
interesting names like ``multi-modal likelihoods and ridges etc...''
The workaround here is to assume that the population is at equilibrium
so that that the x can be replaced with the equilibrium
distribution. However, assuming equilibrium defeats the objective of
many studies of population dynamics, namely determining if and why a
population is at equilibrium.

>>>>

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in
abundance---hypotheses relating to the importance and existence of
phenomenon such as density-dependent population regulation, population
cycling, and spatial synchrony
\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999}.
In applied contexts, population models are used for
estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
In order to address these questions, several complicating factors must
be confronted when fitting population models to data.
First, deterministic models of population
dynamics are virtually always inadequate due to process variation,
the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001}. %clark_bjornstad:2004}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.
%Failure to account for process
%variation and observation error can bias the estimators of abundance
%and related parameters.

State-space models are a widely used approach for
studying population dynamics while accounting for process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Conventional state-space
models are simply time-series models in which the true state of the
system (e.g. population size during each year) is observed
imperfectly. One reason for the widespread adoption of state-space
models in ecology is that failure to account for process variation and
observation error can bias estimators of abundance and population
growth parameters \citep{kery_etal:2009}. For instance, the strength of
density dependence will be over-estimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

One of the simplest state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a deterministic population growth model
(e.g. exponential, logistic, etc \dots) and a stochastic model allowing
for random process variation, which we denote $\grave{o}_t$ . As an
example, under exponential population growth, the model could be
written as:
\begin{gather}
N_t = N_{t-1}\exp(r)+\grave{o}_t \nonumber \\
X_t = N_t + \zeta_t
\label{eq:ss1}
\end{gather}
where $r$ is the intrinsic rate of increase. Typically, both error terms
are assumed to be i.i.d random effects distrubted as
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. The Gaussian assumption allows for parameter
estimation using the Kalman filter \citep{dennis_etal:2006}, which is
much more computationally efficient than estimation methods when
random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}.

Exponential growth without density-dependence is rarely observed in
natural populations, and thus $\exp(r)$ is often substituted for other
population growth models such as the Ricker or Gompertz. Nonetheless,
the basic structure shown in Eq.~\ref{eq:ss1} is common to most
traditional state-space models. However, even though this is
the most widespread approach for modeling population dynamics using
time-series data, several problems are immediately evident. First,
both $N_t$ and $X_t$ are typically non-negative integers, and hence
the the Gaussian assumption is not justified.
%is clearly a problem with this formulation because the normally distributed
%random effects violate this constraint.
One approach to circumventing this problem in hopes of adhering to the
Gaussian assumption is to tranform abundance to
density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
area surveyed. However, this transformation is problematic since
abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
zeros are replaced with some small number, although the
effect of this is rarely discussed.

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Specifically, the Markovian
nature of the model implies the following likelihood
\begin{equation}
\mathcal{L}(\Theta|\{X_t\}) = [N_1|\Theta]\prod_{t=2}^T
[N_t|N_{t-1},\Theta]
\label{eq:L1}
\end{equation}
The first term is the probability density for the first observation in
the time series, $X_1$, yet it is intuitively not possible to estimate the
parameter(s) of this distribution using only a single
observation.
%This problem has resulted in several papers with
%interesting names like ``multi-modal likelihoods and ridges etc...''
Two options have been proposed handling the identifiability
problem. First, one can assume that the first observation in the time
series is made without error. Second, one can assume that the population is at equilibrium
so that that the x can be replaced with the equilibrium
distribution. However, assuming equilibrium defeats the objective of
many studies of population dynamics, namely determining if and why a
population is at equilibrium.
